Viscous Erosion in Porous Media

### 2018-02-22 docs/methods/erosion_methods.tex
NM

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
\label{sec:bies}
Here we describe a boundary integral equation to solve the fluid
equations. In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the grains $\gamma$.  As the viscosity $\mu$
also sets the time scale, we take $\mu=1$.  We start by defining the
double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, $\nn$ is the unit outward
normal at $\yy$, and $\eeta$ is an unknown density function.  The
double-layer potential can not capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}.  The rank deficiency is removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint that results from the flux-free requirement of $\ff$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interface evolution in the {\thL} framework} 
\label{sec:thetaL}

To handle boundary evolution, we use the {\thL} framework, which offers
certain advantages in the ability to stabilize fluid-structure
interaction problems \cite{hou-low-she1994}. Throughout this section, we
use the following conventions: the bodies are parametrized in the
counterclockwise direction and the normal vector points into the bodies
(Fig.~\ref{fig:schematic}).

For the sake of numerical stability, we modify the erosion law in
equation~\eqref{eqn:erosionModel} to include a term that depends on the
local curvature $\kappa$. The idea is to prevent regions of very high
curvature. The modified normal velocity is 
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1
  \right).
\end{align}
Here, $\eps$ is a small parameter, $L$ is the total arc length of the
body, and $\mean{\cdot}$ indicates a boundary mean. The second term on
the right acts to smooth bodies with high-frequency oscillations by
introducing a {\em curvature}-dependent component in the interface
velocity. Since a closed body has a mean curvature of $\mean{\kappa} =
2\pi / L$, the term in parentheses has mean zero. Thus, this term is
area-preserving, so that the only material loss is due to the shear
stress itself. In addition, we have scaled this term on the
$\mean{\atau}$, the average absolute stress, so that it is always
proportional to the true source of erosion $\atau$, even as bodies
vanish. \todo[inline]{reword}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
\label{sec:bies}
Here we describe a boundary integral equation to solve the fluid
equations. In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the grains $\gamma$.  As the viscosity $\mu$
also sets the time scale, we take $\mu=1$.  We start by defining the
double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, $\nn$ is the unit outward
normal at $\yy$, and $\eeta$ is an unknown density function.  The
double-layer potential cannot capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}.  The rank deficiency is removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint associated with the flux-free requirement of $\ff$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary evolution in the {\thL} framework} 
\label{sec:thetaL}

To evolve the boundaries, we use the {\thL} framework, which offers
certain advantages in stabilizing fluid-structure
interaction problems \cite{hou-low-she1994}. Throughout this section, we
use the following conventions: the bodies are parametrized in the
counterclockwise direction and the normal vector points into the bodies
(Fig.~\ref{fig:schematic}).
\todo[inline]{These conventions are probably already stated, right?}

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that depends on the local curvature $\kappa$. The idea is to prevent regions of very high
curvature from developing, as these regions would reduce the accuracy of our fluid solver. We thus modify the normal velocity of each interface as follows
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1
  \right).
\end{align}
The second term on the right is the curvature-dependent modification, the strength of which is controlled by the small parameter $\eps \ll 1$. Here, $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a boundary mean. This term causes regions of high curvature to recede more rapidly, thus smoothing any high-frequency oscillations that my exist. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, this smoothing term has mean zero and and thus preserves area. Therefore, the only source of material loss is the shear-stress itself. In addition, we have scaled the smoothing term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.
